---
title: "Data Analysis Final Project"
output: html_notebook
---

## Loading, Filtering, Inspecting the Data + Binary Classification of Age (young and not young)

```{r}
# Change this depending on where the data is saved on your computer
#file_path <- "~/Downloads/ESS10_CHData.csv" # Noah
file_path <- "C:/Users/phili/Downloads/ESS10_CH Data.csv" # Philipp

ess_data <- read.csv(file_path)

filtered_ess_data <- subset(ess_data, agea <= 100 & trstprt <= 10) # removing nonsense values


filtered_ess_data$agea_binary <- cut(
  filtered_ess_data$agea,
  breaks = c(0, 35, Inf),  # Two bins: <= 35 and > 35 (following with the cutoffs outlined here: https://www.opensocietyfoundations.org/focus/open-society-barometer)
  labels = c(1, 0),  # Labels for binary categories (true for young, false for old)
  right = TRUE  # Include the upper bound in each interval
)

filtered_ess_data$agea_binary <- as.numeric(as.character(filtered_ess_data$agea_binary))
table(filtered_ess_data$agea_binary) # check to see distribution of age once we force it to be binary
```

## Descriptive Statistics - Plotting

```{r}
library(ggplot2)

# Plot frequency distribution for `trstprt`
ggplot(filtered_ess_data, aes(x = trstprt)) +
  geom_histogram(binwidth = 1, fill = "grey", color = "black", alpha = 0.7) +
  labs(title = "Frequency Distribution of Trust in Parliament (trstprt)",
       x = "Trust in Parliament (1-10)",
       y = "Frequency") +
  theme_minimal()

# Plot frequency distribution for `agea`
ggplot(filtered_ess_data, aes(x = agea)) +
  geom_histogram(binwidth = 5, fill = "grey", color = "black", alpha = 0.7) +
  labs(title = "Frequency Distribution of Age (agea_binned)",
       x = "Age",
       y = "Frequency") +
  theme_minimal()
```

## Descriptive Statistics - Central Values

```{r}
median_trstprt <- median(filtered_ess_data$trstprt, na.rm = TRUE)
mean_trstprt <- mean(filtered_ess_data$trstprt, na.rm = TRUE)
mean_age <- mean(filtered_ess_data$agea, na.rm = TRUE)
median_agea <- median(filtered_ess_data$agea, na.rm = TRUE)

# Print results
cat("Trust, Mean / Median", mean_trstprt, " ", median_trstprt, "\n")
cat("Age, Mean / Median", mean_age, " ", median_agea, "\n")
# I think its okay to do either here since we removed outliers and bounded this (see data filtering step)
```

## Descriptive Statistics - SDs + Correlation for Trust and Age (main focal var)

```{r}
sd_trstprt_filtered <- sd(filtered_ess_data$trstprt, na.rm = TRUE)
sd_agea_filtered <- sd(filtered_ess_data$agea, na.rm = TRUE)

cat("Standard Deviation of Trust in Parliament (Filtered):", sd_trstprt_filtered, "\n")
cat("Standard Deviation of Age (Filtered):", sd_agea_filtered, "\n")

pearson_r_age_trust <- cor(filtered_ess_data$agea, filtered_ess_data$trstprt, use = "complete.obs", method = "pearson")
cat("Pearson R for Age & Trust:", pearson_r_age_trust, "\n")

pearson_r_age_binary_trust <- cor(filtered_ess_data$agea_binary, filtered_ess_data$trstprt, use = "complete.obs", method = "pearson")
cat("Pearson R for Age (binary) & Trust:", pearson_r_age_binary_trust, "\n")
```

## Correlation over whole Dataset

```{r}
# This datablock was to explore which variables are associated with trust
# Ensure the necessary libraries are loaded
library(dplyr)

numeric_columns <- sapply(filtered_ess_data, is.numeric)
numeric_data <- filtered_ess_data[, numeric_columns]

# Compute Pearson's R for each column with `trstprt`
correlations <- sapply(numeric_data, function(x) {
  if (!all(is.na(x))) {  # Check if the column is not entirely NA
    cor(x, filtered_ess_data$trstprt, use = "complete.obs", method = "pearson")
  } else {
    NA
  }
})

# Convert the results to a data frame
correlation_df <- data.frame(
  Variable = names(correlations),
  Pearson_R = correlations
)

# Sort the data frame by Pearson's R in descending order
sorted_correlations <- correlation_df %>%
  arrange(desc(Pearson_R))

# Display the sorted correlations
print(sorted_correlations)
```

## Relevant Correlations

```{r}
# Find the Pearson's R between control variables and age
pearson_r_age_happiness <- cor(filtered_ess_data$agea, filtered_ess_data$happy, use = "complete.obs", method = "pearson")
pearson_r_age_happiness
pearson_r_age_covid <- cor(filtered_ess_data$agea, filtered_ess_data$gvhanc19, use = "complete.obs", method = "pearson")
pearson_r_age_covid

pearson_r_age_binary_happiness <- cor(filtered_ess_data$agea_binary, filtered_ess_data$happy, use = "complete.obs", method = "pearson")
pearson_r_age_binary_happiness
pearson_r_age_binary_covid <- cor(filtered_ess_data$agea_binary, filtered_ess_data$gvhanc19, use = "complete.obs", method = "pearson")
pearson_r_age_binary_covid
```

## Relevant Correlation Matrix and Heatmap Visualization

```{r}
# Visualize the relevant correlations in a correlation matrix (based on data preparation from above)
library(ggcorrplot)

vars_of_interest <- c("trstprt", "agea", "agea_binary", "happy", "gvhanc19")
matrix_data <- numeric_data[, intersect(names(numeric_data), vars_of_interest)]
colnames(matrix_data) <- c("Trust", "Happiness", "Age", "COVID_Response", "Age_Binary")

cor_matrix <- cor(matrix_data, use = "complete.obs", method = "pearson")
cor_matrix

ggcorrplot(
  cor_matrix, 
  type = "lower",
  lab = TRUE,
  digits = 4,
  hc.order = FALSE,
  colors = c("blue", "white", "red"),
  title = "Correlation Heatmap"
)
```

## Bivariate Linear Models

```{r}
bivariate_model <- lm(trstprt ~ agea, data = filtered_ess_data)
summary(bivariate_model)

filtered_ess_data$agea_squareroot <- sqrt(filtered_ess_data$agea)
sqrt_model <- lm(trstprt ~ agea_squareroot, data = filtered_ess_data)
summary(sqrt_model)

bivariate_model_binary_age <- lm(trstprt ~ agea_binary, data = filtered_ess_data)
summary(bivariate_model_binary_age)
```

## Multivariate Linear Models

```{r}
sqrt_model_multi <- lm(trstprt ~ agea_squareroot + happy + gvhanc19 + agea_squareroot * gvhanc19, data = filtered_ess_data)
summary(sqrt_model_multi)
binary_model_multi <- lm(trstprt ~ agea_binary + happy + gvhanc19, data = filtered_ess_data)
summary(binary_model_multi)
```

## Point Estimates

```{r}
point_estimate <- data.frame(
  agea_squareroot = c(sqrt(45), sqrt(22)),
  agea_binary = c(0, 1),
  happy = c(5, 7),              
  gvhanc19 = c(2, 4)            
)
predicted_values <- predict(sqrt_model_multi, newdata = point_estimate)
predicted_values
binary_predicted_values <- predict(binary_model_multi, newdata = point_estimate)
binary_predicted_values
```
